Supplementary figures

### 2015-12-15 Manuscript/manuscript.tex
Edits for clarity throughout

\begin{abstract} 
The benefits of high quality habitat for host persistence may be modulated by the presence of an environmentally persistent pathogen. In some cases, the presence of environmental pathogen reservoirs on high quality habitat may lead to the creation of ecological traps, wherein host individuals preferentially colonize high quality habitat, but are then exposed to increased infection risk and disease-induced mortality.  We explored this possibility through the development of a stochastic patch occupancy model, where we varied the pathogen's virulence, transmission rate, and environmental persistence as well as the distribution of habitat quality in the host metapopulation.  This model suggests that for pathogens with intermediate levels of spread, high quality habitat can serve as an ecological trap, and can be detrimental to host persistence relative to low quality habitat.  This inversion of the relative roles of high and low quality habitat highlights the importance of considering the interaction between spatial structure and pathogen transmission when managing wildlife populations exposed to an environmentally persistent pathogen.
\end{abstract}

@@

Since environmental transmission is spatially explicit (i.e. environmental reservoirs can only infect local residents), its role in disease dynamics may further depend on the spatial structure of the host population.  In particular, we expect that host population structure and movement will be influenced by heterogeneity in quality among the habitat patches in a metapopulation.  Indeed, the quality of a habitat patch can affect its extinction and colonization rates, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).  These processes will influence how a pathogen spreads, where environmental pathogen reservoirs get established, and the resulting effects on host occupancy and population size.

We expect that high quality habitat patches, which support greater host density and traffic than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.  As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Robertson2006}, \cite{Almberg2011}).  In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (\cite{Paull2012}).  Similarly, we expect that low quality patches will be less likely to develop pathogen reservoirs and may serve as refuges where susceptible hosts can escape infection.
The combination of these effects may then lead to the creation of metapopulation-scale traps, where the inclusion of otherwise high quality habitat in the metapopulation actually reduces host population size relative to low quality habitat.

@@

The model was parameterized according to the following assumptions (Table~\ref{params}). 
(1) The rate of emigration from a patch increases with the local population size, while the extinction rate of a patch decreases with populations size ($\xi_{em}=0.5$, $\alpha=1$, \cite{Hanski2003}).  
(2) Migrants preferentially select high quality habitat, that is, the rate at which a patch is colonized scales with quality ($\xi_{im}=0.5$, \cite{Hanski2003}, though we relax this below).  
(3)  Patches are arranged in a lattice and only accessible from their four nearest neighbors ($d_{ij}=1$ for all $i$, $j$ neighbors).  
In addition to these assumptions, we chose parameters so that without infection, approximately 0.75 of the patches were occupied ($\mu = 0.1$, $D=2$).  We varied infectious survival ($\nu$), probability of direct transmission ($\delta$), and environmental longevity to explore the dynamics of a range of possible pathogens.

@@

To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat.
These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality, with either the minimum or maximum shifted to simulate regions with more, or less, high quality habitat (yielding distributions with qualities ranging from 0.23 to 1.24 and from 0.76 to 1.77).  
We then implemented a full-factorial set of simulations on each quality distribution in which infectious survival ($\nu$) and probability of direct transmission ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).  In addition, pathogen longevity ($\beta$) was varied over 3 values (ranging from 0.32 to 3.2).

For each simulation -- 100 replicates for each combination of disease parameters ($\nu$, $\delta$, and $\beta$), habitat quality distribution (high or low), and landscape structure (lattice or fully connected), qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above). In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.  The state of the metapopulation was then tracked for 5000 time steps.  We recorded the proportion of patches in each state at the end of each simulation and also computed the mean total population size of susceptible and infectious hosts over the last 500 time steps of the simulation (summing the local population sizes for all occupied patches, assuming a local population size of $A_i$ for patches that are occupied and susceptible and $\nu A_i$ for occupied infectious patches).

To more carefully explore the role of habitat quality in within-patch dynamics, we performed an additional simulation experiment on metapopulations with patch qualities evenly spaced over the full range explored above (from 0.2 to 1.8).
In these simulations, we varied longevity over the values used above and explored the consequences of host preference for high quality habitat (and the corresponding potential for ecological traps) by choosing two different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).  
To isolate the effects of habitat preference and pathogen longevity, we fixed the direct transmission probability ($\delta = 0.5$), infectious survival ($\nu = 0.2$), and landscape structure (lattice).
We performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by susceptible and infectious hosts, and the final total host population size.

@@

A wide range of disease dynamics were observed over the range of parameters and habitat quality distributions explored.  
Epidemic outcome was influenced most strongly by infectious survival, with high infectious survival (i.e. low disease-induced mortality) facilitating widespread infection (in which no susceptible host populations persist), and very low infectious survival (i.e. high disease-induced mortality) facilitating disease-free dynamics or intermediate spread (Fig~\ref{endemic}). 
Pathogens with higher environmental longevities were generally able to invade and spread more easily, even with low infectious survival (Fig~\ref{endemic}(c, f)).  
The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics.  Increasing overall habitat quality facilitated widespread infection for a larger range of parameters (Fig~\ref{endemic}, a-c vs. d-f).
Metapopulations with a fully-connected structure exhibited qualitatively similar patterns, but with generally easier pathogen spread (Supplement).

@@

However, for pathogens falling between these extremes (e.g. with intermediate infectious survival), low quality habitat was a net benefit to the host, supporting larger total population sizes than high quality habitat (blue regions of Fig ~\ref{highvlow}).
The spread of these pathogens was strongly influenced by the habitat quality distribution, with low quality habitat limiting disease spread and better maintaining susceptible occupancy relative to high quality habitat (Fig~\ref{popsizes}, Supplemental).
In these cases, high quality habitat served as a sort of metapopulation-level trap, that is, otherwise beneficial habitat that is detrimental to the overall host population in the presence of disease.
As pathogen longevity increased, the range of the parameter space where we observed this metapopulation-level trap shifted towards more virulent pathogens.
Again, metapopulations with a fully-connected structure displayed similar patterns (Supplement).

@@

This work demonstrates that the habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population.
In some cases, the presence of high quality habitat can enhance the spread and impact of environmentally persistent pathogens.  
In addition, these patches with many resources may become metapopulation-scale ecological traps that are normally beneficial and attractive to hosts, but become a net drain on the metapopulation due to the impacts of disease. 
This suggests that the presence of a disease can lead to a trade-off between otherwise high quality habitat that facilitates metapopulation stability at the expense of increased disease spread, and low quality habitat that hinders disease spread at the expense of metapopulation stability.

This expands on previous studies of disease spread in a homogeneous metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).  
Shifting the patch quality distribution towards high quality habitat in our model is conceptually similar to increasing movement in these models.
And, as in these models, the net consequences of this trade-off for the host metapopulation depend on the characteristics of the disease.
Broadly, we can divide the observed dynamics into three different scenarios: limited, intermediate, and widespread infection (possibly leading to host extinction). 
High quality habitats are a net benefit to the total host population when pathogen spread is either limited or is high enough that both high and low quality patches are affected by disease. 
In these cases, the connectivity and stability of high quality patches facilitate larger total population sizes with little additional pathogen spread.

@@

These results have general management implications that highlight the potential importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.
It is critical to understand these implications in order to avoid falling into the metapopulation trap created by high quality habitat and its interaction with pathogen spread.
Previous theoretical work suggests that management should focus on maintaining patches where conditions are most favourable for the host (i.e. high quality habitat), but Strasser et al. (\cite{Strasser2010}) show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
Ovaskainen and Hanski (\cite{Ovaskainen2003a}) similarly demonstrate that correlated extinctions reduce the contributions of well-connected habitat to metapopulation persistence and increase the contributions of more isolated habitat.
Our model adds to this literature by demonstrating the potential for environmentally persistent pathogens to provide the kind of disturbance under which low quality habitat is beneficial, and thus should be conserved.

@@

\begin{figure}
\includegraphics[angle=270, scale=0.5]{figure/figure_2.eps}
\centering
\caption{Differences between the median total effective population size of hosts on low quality patch distributions (quality between 0.23 to 1.24) and high quality patch distributions (quality between 0.76 and 1.77) as a function of infectious survival ($\nu$) and the probability of direct transmission ($\delta$).  Reds indicate that high quality habitat distributions support higher median host population sizes than low quality habitat, primarily corresponding to pathogens with either very limited or very extensive spread. Conversely, blues indicate higher median host population sizes on low quality habitat distributions, corresponding to pathogens with intermediate levels of spread.  Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.}
\label{highvlow}
\end{figure}

@@

\begin{figure}
\includegraphics[angle=270, scale=0.5]{figure/figure_4.eps}
\centering
\caption{Results from a single representative simulation with habitat quality uniformly distributed between 0.23 and 1.77, and a pathogen with an infectious survival of $\nu = 0.2$, a direct transmission rate of $\delta = 0.5$, and an envioronmental longevity of $\beta = 3.2$.  The top panel shows total proportion of patches occupied by susceptible (blue) and infectious (red) hosts over time.  The bottom panel shows the state -- susceptible (blue), infectious (red), or unoccupied (white) -- of individual patches through time.  Patches are stacked vertically with the lowest quality at the bottom and the highest quality at the top.  Note the overall more consistent occupancy on high quality habitat and the only very brief spurts of infectious occupancy on low quality habitat.}
\label{simvis}
\end{figure}

>>>>

\begin{abstract} 
Habitat quality plays an important role in the dynamics and stability of wildlife metapopulations.
However, the benefits of high quality habitat may be modulated by the presence of an environmentally persistent pathogen. In some cases, the presence of environmental pathogen reservoirs on high quality habitat may lead to the creation of ecological traps, wherein host individuals preferentially colonize high quality habitat, but are then exposed to increased infection risk and disease-induced mortality.  We explored this possibility through the development of a stochastic patch occupancy model, where we varied the pathogen's virulence, transmission rate, and environmental persistence as well as the distribution of habitat quality in the host metapopulation.  This model suggests that for pathogens with intermediate levels of spread, high quality habitat can serve as an ecological trap, and can be detrimental to host persistence relative to low quality habitat.  This inversion of the relative roles of high and low quality habitat highlights the importance of considering the interaction between spatial structure and pathogen transmission when managing wildlife populations exposed to an environmentally persistent pathogen.
\end{abstract}

@@

Since environmental transmission is necessarily a spatial process, its role in disease dynamics may further depend on the spatial structure of the host population.
Many of the hosts listed above have populations that are spread across small patches of suitable habitat in a highly fragmented landscape.
Taking a metapopulation perspective then, we expect that host population structure and movement will be influenced by heterogeneity in the quality of these habitat patches.  Indeed, the quality of a habitat patch can affect its extinction and colonization rates, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).  These processes will influence how a pathogen spreads, where environmental pathogen reservoirs get established, and the resulting effects on host occupancy and population size.

We expect that high quality habitat patches, which are more attractive and support greater host density than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.  As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Robertson2006}, \cite{Almberg2011}).  In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (\cite{Paull2012}).  Similarly, we expect that low quality patches will be less likely to develop pathogen reservoirs and may serve as refuges where susceptible hosts can escape infection.
The combination of these effects may then lead to the creation of metapopulation-scale traps, where the inclusion of otherwise high quality habitat in the metapopulation actually reduces host population size relative to low quality habitat.

@@

The model was parameterized according to the following assumptions (Table~\ref{params}). 
(1) The rate of emigration from a patch increases with the local population size, while the extinction rate of a patch decreases with populations size ($\xi_{em}=0.5$, $\alpha=1$, \cite{Hanski2003}).  
(2) Migrants preferentially select high quality habitat, that is, the rate at which a patch is colonized scales with quality ($\xi_{im}=0.5$, \cite{Hanski2003}).  
(3)  Patches are arranged in a lattice and only accessible from their four nearest neighbors ($d_{ij}=1$ for all $i$, $j$ neighbors).  
In addition to these assumptions, we chose parameters so that without infection, approximately 0.75 of the patches were occupied ($\mu = 0.1$, $D=2$).  We varied infectious survival ($\nu$), probability of direct transmission ($\delta$), and environmental longevity to explore the dynamics of a range of possible pathogens.

@@

To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat.
These two distributions were chosen to explore a possible conservation scenario in which managers might need to prioritize habitat for preservation or restoration.
To capture the consequences of shifting priorities to either low or high quality habitat, we generated patch quality from uniform distributions with equal variance and limits shifted to favor either low or high quality habitat (yielding distributions with qualities ranging from 0.23 to 1.24 and from 0.76 to 1.77).  
We then implemented a full-factorial set of simulations on each quality distribution in which infectious survival ($\nu$) and probability of direct transmission ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).  In addition, pathogen longevity ($\beta$) was varied over 3 values (ranging from 0.32 to 3.2).

For each combination of disease parameters ($\nu$, $\delta$, and $\beta$), habitat quality distribution (high or low), and landscape structure (lattice or fully connected), we ran 100 replicate simulations in which qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above). In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.  The state of the metapopulation was then tracked for 5000 time steps.  We recorded the proportion of patches in each state at the end of each simulation and also computed the mean total population size of susceptible and infectious hosts over the last 500 time steps of the simulation (summing the local population sizes for all occupied patches, assuming a local population size of $A_i$ for patches that are occupied and susceptible and $\nu A_i$ for occupied infectious patches).

To more carefully explore the role of habitat quality in the occupancy of a given patch, we performed an additional simulation experiment on metapopulations with patch qualities evenly spaced over the full range explored above (from 0.2 to 1.8).
In these simulations, we varied longevity over the values used above and explored the consequences of host preference for high quality habitat (and the corresponding potential for ecological traps) by choosing two different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).  
To isolate the effects of habitat preference and pathogen longevity, we fixed the direct transmission probability ($\delta = 0.5$), infectious survival ($\nu = 0.2$), and landscape structure (lattice).
We performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by susceptible and infectious hosts, and the final total host population size.

@@

A wide range of disease dynamics were observed over the range of parameters and habitat quality distributions explored.  
The ability of the pathogen to spread (and of susceptible hosts to persist) was influenced most strongly by infectious survival, with high infectious survival (i.e. low disease-induced mortality) facilitating widespread infection (in which no susceptible host populations persist), and very low infectious survival (i.e. high disease-induced mortality) facilitating disease-free dynamics or intermediate spread (Fig~\ref{endemic}). 
Pathogens with higher environmental longevities were generally able to invade and spread more easily, even with low infectious survival (Fig~\ref{endemic}(c, f)).  
The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics.  Prioritizing high quality habitat facilitated widespread infection for a larger range of parameters (Fig~\ref{endemic}, a-c vs. d-f).
Metapopulations with a fully-connected structure exhibited qualitatively similar patterns, but with generally easier pathogen spread (Supplement).

@@

However, for pathogens falling between these extremes (e.g. with intermediate infectious survival), favoring low quality habitat was a net benefit to the host, with low quality distributions supporting larger total population sizes than high quality distributions (blue regions of Fig ~\ref{highvlow}).
The spread of these pathogens was strongly influenced by the habitat quality distribution, with low quality habitat limiting disease spread and better maintaining susceptible occupancy relative to high quality habitat (Fig~\ref{popsizes}, Supplemental).
In these cases, high quality habitat served as a sort of metapopulation-level trap, that is, otherwise beneficial habitat that is detrimental to the overall host population in the presence of disease.
As pathogen longevity increased, the range of the parameter space where we observed this metapopulation-level trap shifted towards more virulent pathogens.
Again, metapopulations with a fully-connected structure displayed similar patterns (Supplement).

@@

This work demonstrates that the habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population.
In some cases, prioritizing the preservation or restoration of high quality habitat can enhance the spread and impact of environmentally persistent pathogens.  
In addition, these patches with many resources may become metapopulation-scale ecological traps that are normally beneficial and attractive to hosts, but become a net drain on the metapopulation due to the impacts of disease. 
This suggests that the presence of a disease can lead to a trade-off between otherwise high quality habitat that facilitates metapopulation stability at the expense of increased disease spread, and low quality habitat that hinders disease spread at the expense of metapopulation stability.

The net consequences of this trade-off for the host metapopulation depend on the characteristics of the disease.
Broadly, we can divide the observed dynamics into three different scenarios: limited, intermediate, and widespread infection (possibly leading to host extinction). 
High quality habitats are a net benefit to the total host population when pathogen spread is either limited or is high enough that both high and low quality patches are affected by disease. 
In these cases, the connectivity and stability of high quality patches facilitate larger total population sizes with little additional pathogen spread.

@@

These results have general management implications that highlight the potential importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.
It is critical to understand these implications in order to avoid falling into the metapopulation trap created by high quality habitat and its interaction with pathogen spread.
Previous theoretical work suggests that management should focus on maintaining patches where conditions are most favourable for the host (i.e. high quality habitat), but Strasser et al. (\cite{Strasser2010}) show that stochastic disturbance (e.g. disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
Ovaskainen and Hanski (\cite{Ovaskainen2003a}) similarly demonstrate that correlated extinctions reduce the contributions of well-connected habitat to metapopulation persistence and increase the contributions of more isolated habitat.
Our model adds to this literature by demonstrating the potential for environmentally persistent pathogens to provide the kind of disturbance under which low quality habitat is beneficial, and should thus be conserved.

@@

\begin{figure}
\includegraphics[angle=270, scale=0.5]{figure/figure_2.eps}
\centering
\caption{Differences between the median total population size of hosts on low quality patch distributions (quality between 0.23 to 1.24) and high quality patch distributions (quality between 0.76 and 1.77) as a function of infectious survival ($\nu$) and the probability of direct transmission ($\delta$).  Reds indicate that high quality habitat distributions support higher median host population sizes than low quality habitat, primarily corresponding to pathogens with either very limited or very extensive spread. Conversely, blues indicate higher median host population sizes on low quality habitat distributions, corresponding to pathogens with intermediate levels of spread.  Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.}
\label{highvlow}
\end{figure}

@@

\begin{figure}
\includegraphics[angle=270, scale=0.5]{figure/figure_4.eps}
\centering
\caption{Results from a single representative simulation with habitat quality uniformly distributed between 0.23 and 1.77, and a pathogen with an infectious survival of $\nu = 0.2$, a direct transmission rate of $\delta = 0.5$, and an environmental longevity of $\beta = 3.2$.  The top panel shows total proportion of patches occupied by susceptible (blue) and infectious (red) hosts over time.  The bottom panel shows the state -- susceptible (blue), infectious (red), or unoccupied (white) -- of individual patches through time.  Patches are stacked vertically with the lowest quality at the bottom and the highest quality at the top.  Note the overall more consistent infectious occupancy on high quality habitat and the only very brief spurts of infectious occupancy on low quality habitat.}
\label{simvis}
\end{figure}

### 2015-12-15 Manuscript/supplementary.tex
Edits for clarity throughout

\begin{figure}
\includegraphics[angle = 270, scale=0.5]{figure/supplement_2.eps}
\centering
\caption{Differences between the median total effective population size of hosts on high quality patch distributions (quality between 0.76 and 1.77) and low quality patch distributions (quality between 0.23 to 1.24) as a function of infectious survival ($\nu$) and the probability of direct transmission ($\delta$).  Reds indicate higher median host population sizes in high quality patch distributions, while blues indicate higher median host population sizes in low quality patch distributions.  Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.}
\label{highvlow}
\end{figure}

>>>>

\begin{figure}
\includegraphics[angle = 270, scale=0.5]{figure/supplement_2.eps}
\centering
\caption{Differences between the median total population size of hosts on high quality patch distributions (quality between 0.76 and 1.77) and low quality patch distributions (quality between 0.23 to 1.24) as a function of infectious survival ($\nu$) and the probability of direct transmission ($\delta$).  Reds indicate higher median host population sizes in high quality patch distributions, while blues indicate higher median host population sizes in low quality patch distributions.  Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.}
\label{highvlow}
\end{figure}
